---
title: "Customer_Churn"
author: "Dooinn KIM"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
library(tidyverse)
library(lubridate)
library(broom)
library(survival)
library(janitor)
library(ggplot2)
library(survminer)
library(dplyr)


```

## Data import
Dataset avaialbe @ https://www.kaggle.com/datasets/muhammadshahidazeem/customer-churn-dataset

```{r}
df <- read_csv("data.csv")
```
```{r}
df
```
**Variables**:

- ID: CustomerID (identifier, not used in modeling).
- Demographics: Age, Gender.
- Behavioral: Usage Frequency, Support Calls, Payment Delay, Total Spend, Last Interaction.
- Contractual: Tenure, Subscription Type, Contract Length.
- Outcome: Churn (0 = active, 1 = churned).

```{r}
str(df)
summary(df)
colSums(is.na(df))

```
```{r}
df <- na.omit(df)
df <- df %>% clean_names(case = "snake")
```


```{r}
# We mutate to factors so that R understands these variables as categorical, creates the right dummy variables, and gives hazard ratios that make sense.
df <- df %>% mutate(
  gender = as.factor(gender),
  subscription_type = as.factor(subscription_type),
  contract_length = as.factor(contract_length)
)
```

# Churn Distribution

- 0 = Active (censored)
- 1= Churned (event)

- 43% of customers are still active
- 56 % have churned

```{r}

# Frequency bar plot
barplot(table(df$churn), 
        col = c("steelblue", "tomato"),
        main = "Churn Frequency",
        ylab = "Count")

table(df$churn)
prop.table(table(df$churn))

```


## Distibution of Tenure

**1. Summary**:

- Median = 32 months → half of customers churn/are censored before ~2.5 years, half after.
- Distribution is fairly uniform across months, with a slight peak at month 1.

**2. Histogram of tenure**:

- Big spike at 1 month → many customers churn or are censored almost immediately (early drop-off).
- After that, tenure is roughly evenly spread between 2 and 60 months.
- This suggests early disengagement is a major issue (lots of customers don’t “stick” after the first month).

**3. Boxplot: Tenure vs Churn**

- Median tenure for churned customers (Churn = 1) is lower than for active customers.
- Active customers: median ~32 months.
- Churned customers: median closer to ~29 months.
- Distribution spreads overlap a lot, but churners are slightly skewed toward shorter tenure.
- This indicates customers who churn tend to leave earlier than those who stay.

**Key insights**

- Early churn (at month 1) is a critical business issue. Many customers leave right away.
- Churners generally have shorter tenure than retained customers, though the difference is not massive — churn happens across the full range of tenure.
- This pattern validates using survival analysis: the timing of churn is clearly informative, not just whether churn happens.

```{r}
summary(df$tenure)

hist(df$tenure, breaks = 50, col = "skyblue",
     main = "Distribution of Tenure",
     xlab = "Tenure (months)")

boxplot(tenure ~ churn, data = df,
        main = "Tenure vs Churn",
        xlab = "Churn (0=Active, 1=Churned)",
        ylab = "Tenure (months)",
        col = c("lightgreen", "salmon"))

active <- df$tenure[df$churn == 0]
churned <- df$tenure[df$churn == 1]

par(mfrow = c(1, 2))

hist(active, breaks = 50, col = "lightgreen", freq = FALSE,
     main = "Tenure Distribution - Active",
     xlab = "Tenure (months)", xlim = c(0, 60))
lines(density(active), col = "darkgreen", lwd = 2)

hist(churned, breaks = 50, col = "salmon", freq = FALSE,
     main = "Tenure Distribution - Churned",
     xlab = "Tenure (months)", xlim = c(0, 60))
lines(density(churned), col = "red4", lwd = 2)

par(mfrow = c(1, 1))

```

## Descriptive Analysis of Categorical Predictors.

** 1. Gender **

Counts:
- Female = 190,580 (43%)
- Male = 250,252 (57%)

Churn distribution:
- Females: 66.7% churned, 33.3% active
- Males: 49.1% churned, 50.9% active

Interpretation:
- Females churn at a much higher rate than males.
- Males are more likely to remain active customers.
- Gender is likely an important predictor of churn.

** 2. Subscription Type ** 

Counts:
- Basic = 143,026 (32.4%)
- Premium = 148,678 (33.7%)
- Standard = 149,128 (33.8%)

Churn distribution:
- Basic: 58.2% churned, 41.8% active
- Premium: 55.9% churned, 44.1% active
- Standard: 56.1% churned, 43.9% active

Interpretation:
- All three subscription types have very similar churn rates (~56–58%).
- Basic is slightly riskier, but differences are small.
- Subscription type may have limited predictive power compared to gender.


** 3. Contract Length ** 

Counts:
- Annual = 177,198 (40.2%)
- Monthly = 87,104 (19.8%)
- Quarterly = 176,530 (40.0%)

Churn distribution:
- Annual: 46.1% churned, 53.9% active
- Monthly: 100% churned (no active customers!)
- Quarterly: 46.0% churned, 54.0% active

Interpretation:
- Huge insight: Monthly contract customers all churned.
- This makes sense — short-term contracts often have no lock-in, so customers can leave immediately.
- Annual and Quarterly contracts have much better retention (almost identical, ~46% churn).
- Contract length is an extremely strong predictor of churn.


** Summary of categorical predictors **
- Gender: Female churn rate is much higher → strong predictor.
- Subscription type: Similar churn rates across groups → weak predictor.
- Contract length: The most powerful predictor (monthly contracts = guaranteed churn).


```{r}
# Gender
table(df$gender)
prop.table(table(df$gender))

table(df$churn, df$gender)
prop.table(table(df$churn, df$gender), margin = 2) 

barplot(prop.table(table(df$churn, df$gender), margin = 2),
        beside = TRUE, col = c("lightgreen", "salmon"),
        legend = c("Active", "Churned"),
        main = "Churn by Gender")

# Subscription Type
table(df$subscription_type)
prop.table(table(df$subscription_type))

table(df$churn, df$subscription_type)
prop.table(table(df$churn, df$subscription_type), margin = 2)

barplot(prop.table(table(df$churn, df$subscription_type), margin = 2),
        beside = TRUE, col = c("lightgreen", "salmon"),
        legend = c("Active", "Churned"),
        main = "Churn by Subscription Type")


# Contract Length
table(df$contract_length)
prop.table(table(df$contract_length))

table(df$churn, df$contract_length)
prop.table(table(df$churn, df$contract_length), margin = 2)

barplot(prop.table(table(df$churn, df$contract_length), margin = 2),
        beside = TRUE, col = c("lightgreen", "salmon"),
        legend = c("Active", "Churned"),
        main = "Churn by Contract Length")


```


```{r}
# Histograms
hist(df$age, main="Age Distribution", col="lightgreen", xlab="Age")
hist(df$usage_frequency, main="Usage Frequency", col="lightblue", xlab="Usage Frequency")
hist(df$support_calls, main="Support Calls", col="orange", xlab="Support Calls")
hist(df$payment_delay, main="Payment Delay", col="red", xlab="Payment Delay")
hist(df$total_spend, main="Total Spend", col="purple", xlab="Total Spend")

```
## Continuous predictors (age, usage_frequency, support_calls, payment_delay, total_spend, last_interaction) vs churn

** 1. Age **
- Churned customers are older on average (median ~41).
- Active customers are younger (median ~37).
- Older customers appear more likely to churn than younger ones.

** 2.Usage Frequency **
- Active customers have higher usage (median ~16).
- Churned customers use the service less (median ~14).
- Low engagement (lower frequency) is linked to churn risk.

** 3. Support Calls **
- Churned customers contact support more often (median ~5 calls).
- Active customers contact support less (median ~1 call).
- Frequent support calls could reflect problems or dissatisfaction, leading to churn.

** 4. Payment Delay **
- Churned customers have longer delays (median ~15 days).
- Active customers have shorter delays (median ~10 days).
- Poor payment behavior is a warning signal for churn.

** 5. Total Spend **
- Active customers spend much more (median ~750).
- Churned customers spend less (median ~500).
- Higher spenders are more loyal, while low spenders are more likely to churn.

** 6. Last Interaction **
- Churned customers had their last interaction longer ago (median ~15).
- Active customers interacted more recently (median ~11).
- Recent engagement reduces churn risk — inactivity is a churn predictor.

** Key Insights from Continuous Predictors ** 

- Higher churn risk: older, low-usage, frequent support calls, delayed payments, lower spending, and longer since last interaction.
- Protective factors: higher engagement, on-time payments, higher spending, recent activity.
- This aligns perfectly with business intuition: disengaged, frustrated, or low-value customers churn earlier.

```{r}

boxplot(age ~ churn, data=df,
        main="Age vs Churn", col=c("lightgreen","salmon"),
        names=c("Active","Churned"))

boxplot(usage_frequency ~ churn, data=df,
        main="Usage Frequency vs Churn", col=c("lightgreen","salmon"),
        names=c("Active","Churned"))

boxplot(support_calls ~ churn, data=df,
        main="Support Calls vs Churn", col=c("lightgreen","salmon"),
        names=c("Active","Churned"))

boxplot(payment_delay ~ churn, data=df,
        main="Payment Delay vs Churn", col=c("lightgreen","salmon"),
        names=c("Active","Churned"))

boxplot(total_spend ~ churn, data=df,
        main="Total Spend vs Churn", col=c("lightgreen","salmon"),
        names=c("Active","Churned"))

boxplot(last_interaction ~ churn, data=df,
        main="Last Interaction vs Churn", col=c("lightgreen","salmon"),
        names=c("Active","Churned"))

```

** Customer Churn Profile (Descriptive Analysis Summary) **


** Outcome & Time **

Churn rate: 
- ~56% of customers churned, 44% still active.

Tenure:
- Median = 32 months.
- Large spike at 1 month → many customers leave immediately.
- Churners generally have shorter tenure than active customers.



** Categorical Predictors **

Gender:
- Females: 66.7% churned.
- Males: 49.1% churned.
- Females churn more often; being male is protective.

Subscription Type
- Basic, Standard, Premium all have similar churn (~56–58%).
- Subscription type has little impact on churn.

Contract Length
- Monthly: 100% churned (no active customers).

Annual & Quarterly: only ~46% churned.
- Contract length is the strongest predictor: short contracts = guaranteed churn.



** Continuous Predictors **

Age
- Older customers churn more (median ~41 vs ~37).
- Age positively related to churn.

Usage Frequency
- Churners use the service less often.
- Lower engagement = higher churn.

Support Calls
- Churners contact support far more (median ~5 vs ~1).
- Customer frustration strongly linked to churn.

Payment Delay
- Churners delay payments longer (median ~15 days vs ~10).
- Poor payment behavior predicts churn.

Total Spend
- Active customers spend more (median ~750 vs ~500).
- High spenders are more loyal.

Last Interaction
- Churners interacted less recently (median ~15 vs ~11).
- Inactivity predicts churn.


** Curner Profile **

Most likely to churn:

- Female
- On a monthly contract
- Older age
- Lower usage frequency
- More support calls
- Longer payment delays
- Lower spend
- Longer since last interaction

Most likely to stay:

- Male
- On annual or quarterly contracts
- Younger
- Higher engagement (usage frequency)
- Fewer support calls
- Pay on time
- Higher spending
- Recent interaction


## Kaplan–Meier (Overall Survival)
```{r}
library(survival)
library(survminer)

surv_obj <- Surv(time = df$tenure, event = df$churn)

km_fit <- survfit(surv_obj ~ 1, data = df)

summary(km_fit)

ggsurvplot(
  km_fit, 
  conf.int = TRUE, 
  risk.table = TRUE,
  surv.median.line = "hv",
  title = "Overall Customer Survival Curve",
  xlab = "Tenure (months)", 
  ylab = "Survival Probability"
)
```
** Interpretation **

Starting point
- At tenure = 0, survival probability = 1.0 (100% of customers are active at the start).

Decline over time
- The curve steadily decreases as tenure increases.
- Around 30 months, survival probability is ~0.65 → meaning 65% of customers remain active at 2.5 years.
- Around 60 months, survival probability is close to 0 → meaning almost all customers have churned by 5 years.

Early churn
- The curve shows a noticeable steeper drop in the early months, confirming what we saw in the tenure histogram — many customers leave early (especially at month 1).

Number at risk table
- Shows how many customers are still under observation at different time points.
- Starts with 440,832, and by month 60 only 7,658 are left at risk.
- This demonstrates both churn events and right-censoring.

Key Takeaways
- Customer survival probability drops quickly in the first year, then declines steadily.
- Long-term retention is difficult: very few customers survive past 5 years.
- Confirms the importance of early customer engagement and retention strategies.





## Kaplan–Meier (Gender)

```{r}
km_gender <- survfit(surv_obj ~ gender, data = df)

summary(km_gender)

ggsurvplot(
  km_gender,
  conf.int = TRUE,
  risk.table = TRUE,
  pval = TRUE,
  title = "Survival Curves by Gender",
  xlab = "Tenure (months)",
  ylab = "Survival Probability",
  legend.labs = c("Female", "Male"),
  palette = c("salmon", "skyblue"),
  risk.table.height = 0.35,     
  risk.table.y.text.col = TRUE, 
  risk.table.y.text = FALSE     
)


```

** Interpretation of Gender Survival Curves **

Curve shapes
- Both curves decline gradually over time, but female survival is consistently lower than male survival.
- The gap becomes more noticeable after ~20 months and widens as tenure increases.

Survival probabilities
At 30 months:
- Male survival ≈ 0.70
- Female survival ≈ 0.63

At 60 months:
- Male survival ≈ 0.20
- Female survival ≈ 0.05

This means males are more likely to remain active customers at every tenure point.


Statistical test
- p-value < 0.0001 (log-rank test) → the difference between male and female survival is statistically significant.
- So the observed difference is not random noise — gender truly impacts survival.

Number at risk table
- Confirms the dataset size: ~190K females, ~250K males at start.
- Even at 60 months, males retain ~4,367 active customers vs only ~3,291 females.
- This reinforces that male customers survive longer on average.



Key Takeaway
- Gender matters: Female customers churn faster, while male customers show higher long-term retention.
- This confirms what we saw in descriptive stats and boxplots.
- Gender should definitely be included in the survival model as a predictor.



## Kaplan–Meier  (Contract Length)
```{r}

km_contract <- survfit(surv_obj ~ contract_length, data = df)

summary(km_contract)

ggsurvplot(
  km_contract,
  conf.int = TRUE,
  risk.table = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  title = "Survival Curves by Contract Length",
  xlab = "Tenure (months)",
  ylab = "Survival Probability",
  legend.labs = c("Annual", "Monthly", "Quarterly"),
  palette = c("darkgreen", "red", "blue"),
  risk.table.height = 0.35,
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)

```
** Interpretation of Contract Length Survival Curves **

Monthly contracts (red curve)
- Survival declines much faster than Annual/Quarterly.
- By ~30 months, survival probability is already below 0.5.
- By 60 months, almost no monthly contract customers remain.
- Confirms short contracts = high churn.


Quarterly (blue) vs Annual (green)
- Both decline more slowly and have much higher survival than monthly.
- Annual and Quarterly are quite similar — survival ~0.5 at 60 months.
- Long-term contracts protect against churn.


Statistical test
- p < 0.0001 → survival differences across contract lengths are highly significant.


- Number at risk table
- Starts with: Annual ~177K, Monthly ~87K, Quarterly ~176K.
- By 60 months, only a few thousand remain in each group.
- The drop-off for monthly is steepest, while annual & quarterly retain far more customers.


Key Takeaway

- Contract length is the single strongest predictor of survival.
- Customers on monthly contracts churn much earlier than those on annual or quarterly contracts.
- Annual and Quarterly contracts extend customer lifetime, with survival probabilities about 2x higher than monthly by the end of the observation period.


## Kaplan–Meier  (Subscription type)

```{r}
km_subs <- survfit(surv_obj ~ subscription_type, data = df)

ggsurvplot(
  km_subs,
  conf.int = TRUE,
  risk.table = TRUE,
  pval = TRUE,
  surv.median.line = "hv",
  title = "Survival Curves by Subscription Type",
  xlab = "Tenure (months)",
  ylab = "Survival Probability",
  legend.labs = c("Basic", "Premium", "Standard"),
  palette = c("orange", "purple", "blue"),
  risk.table.height = 0.35,
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)

summary(km_subs)

```
** Interpretation of Subscription Type Survival Curves **

Curve shapes
- The three curves (Basic, Premium, Standard) almost completely overlap.
- No group shows a noticeably faster or slower decline.

Survival probabilities
- At 30 months, all three are ~0.65 survival probability.
- At 60 months, all three are close to 0.10–0.15 survival probability.
- No meaningful difference in long-term retention between subscription types.

Statistical test
- p = 0.63 → not statistically significant.
- Confirms what descriptive stats already suggested: subscription type has little or no effect on churn timing.

Number at risk table
- Each subscription type starts with ~143–149K customers.
- By 60 months, only ~2,500–2,600 remain in each group.
- The proportions are virtually the same across all types.

Key Takeaway
- Subscription type does not influence churn survival.
- Basic, Standard, and Premium customers all churn at very similar rates.
- From a modeling standpoint, subscription type may not add much predictive value.




So far with Kaplan–Meier:

- Gender: Significant difference (males survive longer).
- Contract length: Very strong difference (monthly churns fastest).
- Subscription type: No significant difference.





## Nelson–Aalen Cumulative Hazard (Overall)

- Estimates the cumulative hazard function H(t) instead of survival.
- Hazard function describes the instantaneous risk of churn at time t.
- Nelson–Aalen is often plotted alongside KM.

```{r}
km_fit <- survfit(surv_obj ~ 1, data = df)


ggsurvplot(
  km_fit,
  fun = "cumhaz",                
  conf.int = TRUE,              
  risk.table = FALSE,             
  title = "Nelson–Aalen Cumulative Hazard Curve",
  xlab = "Tenure (months)",
  ylab = "Cumulative Hazard"
)

fit_na_gender <- survfit(surv_obj ~ gender, data = df)

ggsurvplot(
  fit_na_gender,
  fun = "cumhaz", 
  conf.int = TRUE,
  risk.table = FALSE,
  title = "Nelson–Aalen Cumulative Hazard by Gender",
  xlab = "Tenure (months)", ylab = "Cumulative Hazard",
  legend.labs = c("Female", "Male"),
  palette = c("salmon", "skyblue")
)

fit_na_contract <- survfit(surv_obj ~ contract_length, data = df)

ggsurvplot(
  fit_na_contract,
  fun = "cumhaz", 
  conf.int = TRUE,
  risk.table = FALSE,
  title = "Nelson–Aalen Cumulative Hazard by Contract Length",
  xlab = "Tenure (months)", ylab = "Cumulative Hazard",
  legend.labs = c("Annual", "Monthly", "Quarterly"),
  palette = c("darkgreen", "red", "blue")
)

fit_na_subs <- survfit(surv_obj ~ subscription_type, data = df)

ggsurvplot(
  fit_na_subs,
  fun = "cumhaz", 
  conf.int = TRUE,
  risk.table = FALSE,
  title = "Nelson–Aalen Cumulative Hazard by Subscription Type",
  xlab = "Tenure (months)", ylab = "Cumulative Hazard",
  legend.labs = c("Basic", "Premium", "Standard"),
  palette = c("orange", "purple", "blue")
)
```

## Cumulative Incidence of Churn

- Cumulative incidence = probability that an event (churn) has occurred by a certain time t. 
- In other words, instead of asking “what % of customers are still active at 30 months?”, we ask “what % have churned by 30 months?”.

```{r}

km_fit <- survfit(surv_obj ~ 1, data = df)

ggsurvplot(
  km_fit,
  fun = "event",  
  conf.int = TRUE,
  risk.table = TRUE,
  title = "Cumulative Incidence of Churn (1 - Survival)",
  xlab = "Tenure (months)",
  ylab = "Cumulative Probability of Churn"
)

```
Start point
- At tenure = 0, cumulative incidence = 0 (no one churned yet).

Steady rise
- The curve increases gradually → showing more customers churn over time.
- At ~30 months: cumulative incidence ≈ 35% → about one-third of customers have churned by 2.5 years.
- At ~50 months: cumulative incidence ≈ 70% → about 7 in 10 customers have churned.
- At ~60 months: cumulative incidence ≈ 95%+ → almost all customers churn within 5 years.

Curve shape
- The curve bends upward in later months → churn accelerates as customers stay longer.
- This matches the Nelson–Aalen hazard curve, which showed risk piling up after ~40 months.

Key Takeaway
- Cumulative incidence of churn = probability of churn by time t.
- “What % of my customers will have left by month X?”
- By 2.5 years → ~1 in 3 customers churn.
- By 5 years → nearly all customers churn.



```{r}

# 1. By Gender
fit_cif_gender <- survfit(surv_obj ~ gender, data = df)
ggsurvplot(
  fit_cif_gender,
  fun = "event",
  conf.int = TRUE,
  risk.table = TRUE,
  pval = TRUE,
  title = "Cumulative Incidence of Churn by Gender",
  xlab = "Tenure (months)", ylab = "Cumulative Probability of Churn",
  legend.labs = c("Female", "Male"),
  palette = c("salmon", "skyblue"),
  risk.table.height = 0.35
)

# 2. By Contract Length
fit_cif_contract <- survfit(surv_obj ~ contract_length, data = df)
ggsurvplot(
  fit_cif_contract,
  fun = "event",
  conf.int = TRUE,
  risk.table = TRUE,
  pval = TRUE,
  title = "Cumulative Incidence of Churn by Contract Length",
  xlab = "Tenure (months)", ylab = "Cumulative Probability of Churn",
  legend.labs = c("Annual", "Monthly", "Quarterly"),
  palette = c("darkgreen", "red", "blue"),
  risk.table.height = 0.35
)

# 3. By Subscription Type
fit_cif_subs <- survfit(surv_obj ~ subscription_type, data = df)
ggsurvplot(
  fit_cif_subs,
  fun = "event",
  conf.int = TRUE,
  risk.table = TRUE,
  pval = TRUE,
  title = "Cumulative Incidence of Churn by Subscription Type",
  xlab = "Tenure (months)", ylab = "Cumulative Probability of Churn",
  legend.labs = c("Basic", "Premium", "Standard"),
  palette = c("orange", "purple", "blue"),
  risk.table.height = 0.35
)



```

## Overall Hazard Function

Among customers who have not yet churned at month t, what is the probability they will churn in the next instant (e.g., next month)?

```{r}
fit <- survfit(Surv(tenure, churn) ~ 1, data = df)

haz_df <- data.frame(
  time = fit$time,
  cumhaz = -log(fit$surv)
)

haz_df$hazard <- c(haz_df$cumhaz[1], diff(haz_df$cumhaz))

plot(haz_df$time, haz_df$hazard, type="l", col="blue", lwd=2,
     main="Estimated Hazard Function of Churn",
     xlab="Tenure (months)", ylab="Hazard")

```
Early months (0–20)
- Hazard is very low (close to 0).
- if a customer survives the very first months, their risk of churning remains modest.

Middle period (20–40 months)
- Hazard increases slightly but remains relatively low.
- Risk of churn is steady and manageable.

Later period (40–60 months)
- Hazard grows exponentially.
- At ~60 months, the hazard spikes → meaning customers who last this long are very likely to churn soon.
- Matches what we saw in cumulative incidence: almost all customers churn by 5 years.



** Role of the Hazard Function **

- Complements survival probability

Survival function 
- S(t): “How many remain?”

Hazard function 
- h(t): “How risky is it right now?”

Reveals churn dynamics
- Early vs late risk: tells you when churn is most dangerous.
- In the churn analysis, risk is low at the start but accelerates dramatically later.

Inputs for modeling
- In Cox regression, we model hazard ratios between groups (e.g., males vs females, annual vs monthly).
- Helps quantify how predictors shift the risk curve.

Business insight
- Tells you when to intervene:
- Early customers → onboarding & engagement.
- Long-tenure customers → loyalty programs to stop late-stage churn.

In short: The hazard function tells you “when churn risk is highest”, while the survival function tells you “how many customers remain”.


```{r}
groups <- split(df, df$gender)

par(mfrow=c(1,2)) 
for (g in names(groups)) {
  fit <- survfit(Surv(tenure, churn) ~ 1, data = groups[[g]])
  haz_df <- data.frame(
    time = fit$time,
    cumhaz = -log(fit$surv)
  )
  haz_df$hazard <- c(haz_df$cumhaz[1], diff(haz_df$cumhaz))
  
  plot(haz_df$time, haz_df$hazard, type="l", lwd=2,
       main=paste("Hazard Function -", g),
       xlab="Tenure (months)", ylab="Hazard")
}
par(mfrow=c(1,1))


groups <- split(df, df$contract_length)

par(mfrow=c(1,3)) 
for (g in names(groups)) {
  fit <- survfit(Surv(tenure, churn) ~ 1, data = groups[[g]])
  haz_df <- data.frame(
    time = fit$time,
    cumhaz = -log(fit$surv)
  )
  haz_df$hazard <- c(haz_df$cumhaz[1], diff(haz_df$cumhaz))
  
  plot(haz_df$time, haz_df$hazard, type="l", lwd=2,
       main=paste("Hazard Function -", g),
       xlab="Tenure (months)", ylab="Hazard")
}
par(mfrow=c(1,1))


```



## Log-rank Test & Fleming-Harrington Test

Log-rank test and Fleming–Harrington test are non-parametric hypothesis tests used to compare survival curves across groups.

1. Log-rank Test
- Purpose: Tests whether two (or more) survival curves are statistically different.
- Null hypothesis H0 : Survival curves are the same across groups.
- Alternative: At least one group has different survival.

2. Fleming–Harrington Test

- A generalization of the log-rank test.
- The log-rank test gives equal weight to all event times.
- But sometimes we want to give more weight to early events (customers churn quickly) or late events (customers churn after staying long).


When to use?

- Log-rank → default, best when hazard functions are proportional.
- Fleming–Harrington → useful if survival curves cross or if you care about early vs late churn specifically.

```{r}
library(survival)

# Gender (2 groups)
survdiff(Surv(tenure, churn) ~ gender, data = df)       
survdiff(Surv(tenure, churn) ~ gender, data = df, rho=1)  

survdiff(Surv(tenure, churn) ~ contract_length, data = df)       
survdiff(Surv(tenure, churn) ~ contract_length, data = df, rho=1) 

# Subscription type (3 groups)
survdiff(Surv(tenure, churn) ~ subscription_type, data = df)      
survdiff(Surv(tenure, churn) ~ subscription_type, data = df, rho=1) 
```


```{r}
library(survival)
library(survMisc)


ten_gender   <- ten(Surv(tenure, churn) ~ gender,            data = df)
ten_contract <- ten(Surv(tenure, churn) ~ contract_length,   data = df)
ten_subs     <- ten(Surv(tenure, churn) ~ subscription_type, data = df)

# Run FH tests

comp(ten_gender,   p=0, q=0)   # log-rank
comp(ten_gender,   p=1, q=0)   # early-weighted
comp(ten_gender,   p=0, q=1)   # late-weighted

comp(ten_contract, p=0, q=0)
comp(ten_contract, p=1, q=0)
comp(ten_contract, p=0, q=1)

comp(ten_subs,     p=0, q=0)
comp(ten_subs,     p=1, q=0)
comp(ten_subs,     p=0, q=1)

```


Gender
- Log-rank (ρ=0)
- Female: 127,058 observed vs 107,387 expected events (churns).
- Male: 122,941 observed vs 142,612 expected.
- Chisq = 6,595, df=1, p < 2e-16.
- Females churn much more than expected, males churn much less. Highly significant difference.

Weighted (ρ=1, early emphasis)
- Still significant: Chisq = 5,566, df=1, p < 2e-16.
- Confirms females churn earlier and more often than males.
- Gnder strongly affects churn survival, both early and overall.

Contract Length
- Log-rank (ρ=0)
- Annual/Quarterly: fewer churns than expected.
- Monthly: 87,104 observed vs 47,909 expected → way higher churn.
- Chisq = 41,409, df=2, p < 2e-16.

Weighted (ρ=1, early emphasis)
- Monthly still massively worse: 56,963 observed vs 31,512 expected.
- Chisq = 34,900, df=2, p < 2e-16.
- Contract length is the strongest predictor: monthly contracts churn very early, while annual/quarterly are protective.

Subscription Type
- Log-rank (ρ=0)
- Observed ≈ Expected in all groups.
- Chisq = 0.9, df=2, p = 0.6 → not significant.

Weighted (ρ=1)
- Slight imbalance, but still non-significant (Chisq = 4.3, df=2, p = 0.1).
- Subscription type has no meaningful effect on churn timing.


** Summary of Tests **

- Gender: Females churn significantly earlier/more than males. 
- Contract length: Monthly customers churn much earlier and more often. Strongest signal.
- Subscription type: No significant effect.



## Cox Proportional Hazards Modeling

Purpose
- Quantify the effect of predictors on chunrn risk over time.
- Expressed as Hazard Ratios (HRs)
  - HR > 1 : Higher risk of Churn
  - HR < 1 : Lower Risk (Protective)
  - HR = 1 : No effect
  
```{r}
# Base Cox Model
cox_model <- coxph(
  Surv(tenure, churn) ~ gender + contract_length + subscription_type +
    age + usage_frequency + support_calls + payment_delay +
    total_spend + last_interaction,
  data = df
)

summary(cox_model)
```


Categorical Variables

Gender (Male vs Female [ref])
- HR = 0.823, p < 2e-16.
- Men have ~18% lower risk of churn than women, holding everything else constant.
- Females churn faster than males.

Contract Length
- Monthly vs Annual [ref]: HR = 1.622, p < 2e-16. → Monthly contracts have a 62% higher churn risk compared to annual.
- Quarterly vs Annual: HR ≈ 1.00, p = 0.949. → No meaningful difference from annual.
- Monthly is risky, quarterly ≈ annual.

Subscription Type
- Premium: HR = 1.00 (NS).
- Standard: HR = 0.99 (NS).
- No meaningful effect on churn compared to Basic.

Continuous Variables

Age: 
- HR = 1.008 per year, p < 2e-16.
- Each additional year of age increases churn risk by ~0.8%.
- Older customers churn slightly faster.

Usage Frequency:
- HR = 0.999, p < 0.001.
- More usage → slightly lower churn risk.→ Engagement is protective.


Support Calls: 
- HR = 1.118 per call, p < 2e-16.
- Each additional support call increases churn risk by 12%.→ Frequent support users churn much more.

Payment Delay: 
- HR = 1.019 per unit, p < 2e-16.
- Every extra delay increases churn risk by ~2%. → Late payers churn more.

Total Spend
- HR = 0.999, p < 2e-16.
- Higher spend slightly reduces churn risk. → High-spenders are more loyal.

Last Interaction: 
- HR = 1.010, p < 2e-16.
- More recent interactions increase churn risk slightly.→ Suggests that customers reaching out more recently are at higher churn risk (likely dissatisfied).


Model Fit
- Concordance (C-index) = 0.706
- Good discrimination → the model correctly orders churn times ~71% of the time.
- All global tests (LR, Wald, Score) highly significant (p < 2e-16).
- Model explains churn timing very well.

Insights:
- Biggest risks: Monthly contracts, high support callers, frequent late payments.
- Protective factors: Male gender, higher spend, higher usage frequency.
- Neutral factors: Subscription type, quarterly contracts.



## Check Schoenfeld residuals

What are Schoenfeld Residuals?
- In a Cox model, each event (churn) has a residual for every covariate.
- They measure the difference between the observed covariate value at the event time and the expected covariate value given the risk set.
- If the PH assumption holds → the residuals should be uncorrelated with time.
- If PH is violated → the residuals will show a systematic trend with time.

```{r}
# Check Proportional Hazards Assumption
ph_test <- cox.zph(cox_model)
print(ph_test)   # p > 0.05 → assumption holds
plot(ph_test)
```
Results of cox.zph(cox_model)
- Gender:	5.1e-05 /  PH assumption violated → effect of gender on churn changes over time.
- Contract length	; < 2e-16	/ strong violation → hazard ratio for monthly vs annual is not constant across tenure.
- Subscription type	;	0.0019	/ Violation (small effect, but statistically significant).
- Age	;	1.2e-07	/ Effect of age is time-varying.
- Usage frequency	;	1.0e-07	/ Violation.
- Support calls	;	< 2e-16	/ Strong violation.
- Payment delay	;	1.8e-09	/ Violation.
- Total spend	;	< 2e-16	/ Strong violation.
- Last interaction	;	6.5e-07	/ Violation.
- GLOBAL	;	< 2e-16	/ Overall PH assumption does not hold for the model.


The proportional hazards assumption is violated for almost all predictors.
In other words: the hazard ratios are not constant over time.
What we saw in Kaplan–Meier and hazard plots (e.g., monthly contract customers churning early) is confirmed here.

For example:
The effect of monthly contracts might be much stronger in the first year, then weaker later.
The effect of support calls may grow stronger as tenure increases.



## Cox model with stratification by contract length

- contract_length showed the strongest violation of the proportional hazards assumption.
- Stratification removes its hazard ratio estimate but allows each contract type (Annual, Monthly, Quarterly) to have its own baseline hazard curve.
- This way, the effects of other covariates are estimated within each stratum, while contract_length differences are absorbed.

```{r}
# Stratified Cox PH model (stratify by contract_length)
strat_model <- coxph(
  Surv(tenure, churn) ~ gender + subscription_type + age +
    usage_frequency + support_calls + payment_delay +
    total_spend + last_interaction +
    strata(contract_length),
  data = df
)

summary(strat_model)
```


```{r}
cox.zph(strat_model)
```
Interpretation:
- Stratifying by contract length fixed its violation, but others remain (support calls, spend, etc.).
- Hazard ratios give strong, interpretable signals (support calls ↑ risk, total spend ↓ risk).
- For more rigorous modeling, we’ll need time-varying effects or AFT models.




## Accerlerated Failure Time (AFT) model

What is an AFT model?
- Instead of modeling hazard ratios (like Cox), the AFT model assumes that covariates accelerate or decelerate survival time.
- It answers: “How much longer or shorter do customers survive before churn, given their characteristics?”
- The effect is multiplicative on time, not hazard.
- Example: If genderMale = 1.2, men’s survival times are 20% longer than women’s.



Common AFT distributions
 -Exponential (constant hazard, often too simple).
- Weibull (flexible, hazard can increase or decrease).
- Log-normal (time to event follows log-normal).
- Log-logistic (allows hazards to rise then fall).
- We usually compare them via AIC / BIC to see which fits best.


```{r}
# Weibull AFT model
aft_weibull <- survreg(Surv(tenure, churn) ~ gender + contract_length +
                         subscription_type + age + usage_frequency +
                         support_calls + payment_delay + total_spend +
                         last_interaction,
                       data = df,
                       dist = "weibull")

summary(aft_weibull)

# Try log-normal as well
aft_lognormal <- survreg(Surv(tenure, churn) ~ gender + contract_length +
                           subscription_type + age + usage_frequency +
                           support_calls + payment_delay + total_spend +
                           last_interaction,
                         data = df,
                         dist = "lognormal")

summary(aft_lognormal)

# Compare model fit (lower AIC = better)
AIC(aft_weibull, aft_lognormal)

```
Model Fit (Which is Better?)

- Weibull AFT:
  - LogLik = –1,161,223
  - AIC = 2,322,473

- Log-normal AFT:
  - LogLik = –1,201,799
  - AIC = 2,403,623

- Lower AIC is better → WeibullAFT model fits your data better than Log-normal.


Interpreting Weibull AFT Coefficients

- In AFT models, coefficients represent log-time ratios.
- Positive coef → longer survival (slower churn).
- Negative coef → shorter survival (faster churn).
- exp(coef) → time ratio (multiplier for survival time).

Interpretaion on key predictors:
- Gender (Male = +0.114): exp(0.114) ≈ 1.12 → men survive ~12% longer than women before churning.
- Contract length Monthly (–0.279) exp(–0.279) ≈ 0.76 → monthly contracts shorten survival time by ~24% compared to annual (baseline).


Cox gave hazard ratios (risk multipliers).

AFT gives time ratios (survival duration multipliers).

Stick with Weibull AFT as best-fitting parametric model.

Summary

- The Weibull AFT model is best parametric fit.
- Major drivers of shorter survival: monthly contracts, more support calls, higher payment delays, longer since last interaction, older age.
- Protective factors: being male, higher total spend, higher usage frequency.


```{r}
library(survival)


predict_survival <- function(model, newdata, times) {
  lp <- predict(model, newdata = newdata, type = "lp")
  scale <- model$scale
  sapply(times, function(t) exp(- (t / exp(lp))^(1/scale)))
}

times <- seq(1, 60, by = 1)


profiles <- list(
  Profile1 = data.frame(
    gender = "Male",
    contract_length = "Annual",
    subscription_type = "Basic",
    age = 40,
    usage_frequency = 10,
    support_calls = 2,
    payment_delay = 5,
    total_spend = 500,
    last_interaction = 10
  ),
  Profile2 = data.frame(
    gender = "Female",
    contract_length = "Monthly",
    subscription_type = "Premium",
    age = 30,
    usage_frequency = 20,
    support_calls = 5,
    payment_delay = 2,
    total_spend = 1000,
    last_interaction = 3
  ),
  Profile3 = data.frame(
    gender = "Male",
    contract_length = "Quarterly",
    subscription_type = "Standard",
    age = 55,
    usage_frequency = 5,
    support_calls = 1,
    payment_delay = 0,
    total_spend = 300,
    last_interaction = 20
  )
)


surv_profiles <- lapply(profiles, function(p) {
  predict_survival(aft_weibull, newdata = p, times)
})


plot(times, surv_profiles[[1]], type = "l", lwd = 2, col = "blue",
     ylim = c(0,1), xlab = "Tenure (months)", ylab = "Survival Probability",
     main = "Predicted Survival Curves by Different Profiles")
lines(times, surv_profiles[[2]], col = "red", lwd = 2)
lines(times, surv_profiles[[3]], col = "green", lwd = 2)
legend("topright", legend = names(profiles), col = c("blue","red","green"), lwd = 2)
```

Cox Proportional Hazards Model

- Semi-parametric: It doesn’t assume a distribution for survival times.
- Focuses on hazard ratios (relative risk).
- Assumes proportional hazards (the effect of a covariate is constant over time).
- Flexible, widely used, easier to interpret in terms of risk.
- But… in your results, the Schoenfeld test + residual plots showed violations of proportional hazards, which weakens the Cox model’s validity.



Accelerated Failure Time (AFT) Model

- Parametric: Assumes a distribution for survival times (Weibull, lognormal, etc.).
- Focuses on time ratios: covariates accelerate or decelerate the survival time.
- More interpretable in business terms: e.g., “Monthly contracts reduce survival time by 40% compared to Annual.”
- Handles non-proportional hazards better (since it models survival time directly).
- Needs the right distribution — misspecification can bias results.
- In your case, the Weibull AFT had better fit (lower AIC) than lognormal.


Results Comparison
- Cox model: Concordance ~0.70 (pretty good), but proportional hazards assumption failed (cox.zph significant for many covariates).
- AFT (Weibull): AIC much lower than lognormal, strong significance for most predictors, interpretable as “time acceleration/deceleration.”
- Subscription type remained not significant in both models → consistent.


Conclusion
- Since proportional hazards are violated, the AFT model is more appropriate.
- Among AFT options, Weibull fits better than lognormal (based on AIC).
- But Cox can still be used as a benchmark model, since many readers are more familiar with hazard ratios.






